---
title: "Additional Material for Correlation and Regression Chapters"
subtitle: "Additional material for the chapters 'Correlation' and 'Regression' from Breakwell, et al (2025) Research Methods in Psychology"
editor: visual
title-block-banner: true

format: 
  html5:
    fig-pos: "H"
    embed-resources: true
author:
  - name: Jeremy Miles
    email: jeremy.miles@gmail.com*
    affiliations:
      - id: USC
        name: University of Southern California, Los Angeles
        city: Los Angeles, USA
      - id: Google
        name: Google, Los Angeles, USA
        city: Los Angeles, USA
  - name: Mark Shevlin
    email: m.shevlin@ulster.ac.uk
    affiliations:
      - name: University of Ulster, N Ireland, UK
date: today        
license: "GPL 3.0"
---

This document contains links to data, and all of the code to reproduce the examples found in the chapters on Correlation and Regression in Breakwell, et al (2025), Research Methods in Psychology, published by Sage.

This page is written using [Quarto](https://notebooklm.google.com/notebook/1efedbc4-8bc9-4dc2-af1c-466bd097e424/audio), a scientific publishing system that lets you link data and code to produce a document. This document, and the source code that can be used to create this document can be found at <https://github.com/jeremymiles/rmip>. You will need either [RStudio](https://posit.co/download/rstudio-desktop/) or [Positron](https://positron.posit.co/), both of which are free to download, to compile this document.

This document has a [GPL 3.0](https://www.gnu.org/licenses/gpl-3.0.en.html) licence. That means you are free to share, adapt or modify the document, but you also need to share the source code ([the quarto markdown document](https://github.com/jeremymiles/rmip/blob/main/correlation_and_regression.qmd)) for free if you do that.

# (Sort of) Podcasts 

We have used [NotebookLM](https://notebooklm.google.com/) to create two podcasts that cover the material in each chapter. They are generated by AI, and they are not perfect, they add some information that is not in the chapters, but we think do quite a good job of covering the material in the chapters.

-   [Link to correlation 'podcast'](https://notebooklm.google.com/notebook/1efedbc4-8bc9-4dc2-af1c-466bd097e424/audio) (\~26 minutes)

-   [Link to regression 'podcast'](https://notebooklm.google.com/notebook/c7730339-2a09-417e-a0d2-922b3a9c9cd9/audio) (\~37 minutes)\

# Correlation

## The Code 

First we need to load the packages that we're going to be using in the document. If you don't have these installed in your R environment, you need to add them, which you almost certainly do by running:

`install.packages("package_name")`

Note that in the code below, we've used `output=FALSE` in this code block We did that so that the document isn't full of messages about each package. If you run into problems, change that to `TRUE` and you might get a helpful error message.

```{r, output=FALSE}
library(boot.pval)
library(car)
library(datasauRus)
library(datasets)
library(dplyr)
library(ggplot2)
library(haven)
library(lmtest)
library(lubridate)
library(MASS)
library(NHANES)
library(QuantPsyc)
library(sandwich)
library(stringr)

```

### Galton

First we'll load the Galton data. The data are loaded straight from github, so we assume that you have an internet connection to do this.

```{r}
d_galton <- 
  read.table(
    paste0("https://raw.githubusercontent.com/jeremymiles/", 
           "rmip/refs/heads/main/Galton.dat"), 
    header = TRUE)

head(d_galton)
```

We only want to look at the boys, so we will filter the data to include them, and then we can produce the plot in Figure 1.

```{r}
#| label: fig-charts
#| fig-cap: " Scatter plot of heights of fathers and sons from Galton’s data."
d_male <- d_galton %>% 
  dplyr::filter(gender == "M")

# Figure 1
ggplot2::ggplot(d_male, aes(x = father, y = height)) +
  geom_point() +
  ylab("Son's height (inches)") +
  xlab("Father's height (inches)") +
  theme_gray(base_size = 18)

```

We can calculate the correlation in this data.

```{r}
cor(d_male$father, d_male$height)
```

The `echo: false` option disables the printing of code (only output is displayed).
