---
title: "Additional Material for Correlation and Regression Chapters"
subtitle: "Additional material for the chapters 'Correlation' and 'Regression' from Breakwell, et al (2025) Research Methods in Psychology"
editor: visual
title-block-banner: true

format: 
  html5:
    fig-pos: "H"
    embed-resources: true
author:
  - name: Jeremy Miles
    email: jeremy.miles@gmail.com*
    affiliations:
      - id: USC
        name: University of Southern California, Los Angeles
        city: Los Angeles, USA
      - id: Google
        name: Google, Los Angeles, USA
        city: Los Angeles, USA
  - name: Mark Shevlin
    email: m.shevlin@ulster.ac.uk
    affiliations:
      - name: University of Ulster, N Ireland, UK
date: today        
license: "GPL 3.0"
---

This document contains links to data, and all of the code to reproduce the examples found in the chapters on Correlation and Regression in Breakwell, et al (2025), Research Methods in Psychology, published by Sage.

This page is written using [Quarto](https://notebooklm.google.com/notebook/1efedbc4-8bc9-4dc2-af1c-466bd097e424/audio), a scientific publishing system that lets you link data and code to produce a document. This document, and the source code that can be used to create this document can be found at <https://github.com/jeremymiles/rmip>. You will need either [RStudio](https://posit.co/download/rstudio-desktop/) or [Positron](https://positron.posit.co/), both of which are free to download, to compile this document.

This document has a [GPL 3.0](https://www.gnu.org/licenses/gpl-3.0.en.html) licence. That means you are free to share, adapt or modify the document, but you also need to share the source code (the quarto document) for free if you do that.

# (Sort of) Podcasts 

We have used [NotebookLM](https://notebooklm.google.com/) to create two podcasts that cover the material in each chapter. They are generated by AI, and they are not perfect, they add some information that is not in the chapters, but we think do quite a good job of covering the material in the chapters.

-   [Link to correlation 'podcast'](https://notebooklm.google.com/notebook/1efedbc4-8bc9-4dc2-af1c-466bd097e424/audio) (\~26 minutes)

-   [Link to regression 'podcast'](https://notebooklm.google.com/notebook/c7730339-2a09-417e-a0d2-922b3a9c9cd9/audio) (\~37 minutes)

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
